The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-02-18 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revision of discussion regarding linear models
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2468 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 724c1d12ecdff22908887a991eb10f6768d10b9f

From comparing the effects of the building blocks three main conclusions can be drawn: 
(1)~In most cases including all available score values rather than just residuals leads to 
clearly better results and performs in general at least equally well.
(2)~Binarizing scores/residuals does not lead to any improvement in performance.
(3)~Categorizing possible split variables also doesn't improve the performance 
unless the breakpoints used for the categorization are very close to the true split points.
However, even then the results are very similar to those achieved without any categorization.
But it has to be noted that a categorization of possible split variables can facilitate
evaluations. Moreover, depending on the focus of interest, the results can be more
straightforward, for example in case of clearly separated subgroups being defined by the values 
of a continuous split variable. Therefore, categorizing possible split variable can have
advantages in practice, however, it has to be considered that in general
important effects might be neglected.

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
Furthermore, with only two parameters determining a linear model the importance of including scores 
regarding each parameter could already be pointed out clearly. But for an increasing number of 
parameters in a model this aspect becomes even more important as it allows for the detection of changes 
in each parameter.

Note that considering full model scores is particularly advantageous if the model parameters do not
depend on each other, hence are orthogonal, as it is the case for the intercept and slope parameter
of a linear model. On the other hand, if all parameters are closely correlated with the location parameter
a residual-based approach might be able to detect changes in other parameters than the mean as well,
however, the performance of a score-based search for split variables is not negatively effected.

>>>>

From comparing the effects of the building blocks three main conclusions can be drawn: 
(1)~In many cases including all available score values rather than just residuals leads to 
clearly better results and performs in general at least equally well.
(2)~Binarizing scores/residuals does not lead to any improvement in performance.
(3)~Categorizing possible split variables also doesn't improve the performance 
unless the breakpoints used for the categorization are very close to the true split points.
However, even then the results are very similar to those achieved without any categorization.
But it has to be noted that a categorization of possible split variables can facilitate
evaluations. Moreover, depending on the focus of interest, the results can be more
straightforward, for example in case of clearly separated subgroups being defined by the values 
of a continuous split variable. Therefore, categorizing possible split variable can have
advantages in practice, however, it has to be considered that in general
important effects might be neglected.

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
One specific feature of linear models is the orthogonality of the conditional expectation
$\mathbb{E}(Y|X)$ and the regressor $X$ which affects the capability of tests using only residuals
to capture shifts in the parameters. In particular, \cite{Ploberger+Kramer:1992} have shown that
residual-based tests are insensitive to parameter changes orthogonal to the mean regressor.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
Testing procedures purely based on residuals perform best in situations with shifts only in the
intercept, however, the costs in terms of power that come with including all scores are 
negligibly small. 
Furthermore, with only two parameters determining a linear model the importance of including scores 
regarding each parameter could already be pointed out clearly while for an increasing number of 
parameters in a model this aspect becomes even more important.

In general residual-based and score-based procedures are expected to perform equally well if all
parameters are highly correlated. On the other hand, if parameters are orthogonal changes that are 
not connected to the mean might be missed when only considering residuals. This can be avoided by
including scores which are a standard measurement of goodness of fit that can be easily employed.
Even though it also depends on the type of the applied residuals 
(see, e.g., \cite{Loh+Zheng:2013}, \cite{Chaudhuri+Loh:2002} 
and \cite{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made.
